Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-13 paper.tex
dnj edits, new video

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming tools, especially for live programming and end user programming.}
\author{Daniel Jackson}
\authorinfo{is a professor in the Department of Electrical Engineering and Computer Science at MIT, associate director of CSAIL, and a MacVicar Fellow. He leads the Software Design Group.}
\affiliation{Massachusetts Institute of Technology}

@@

We present the idea of \emph{spreadsheet-driven customization}: enabling
end users to customize existing applications using a live spreadsheet
view of the data inside the application. By manipulating the
spreadsheet, users can implement a wide variety of customizations,
ranging from sorting lists of search results to displaying related data
from other web services, without doing any traditional programming.

We built a prototype system called Wildcard that implements
spreadsheet-driven customization as a web browser extension. Through
concrete examples, we demonstrate that Wildcard has both a low barrier
to entry for beginners and enough flexibility to solve many useful
problems. We also show that Wildcard can work with real existing
websites, by extracting structured data using web scraping techniques.

@@

In this paper, we present \emph{spreadsheet-driven customization}, a
technique that enables end users to customize software without doing any
traditional programming. The idea is to augment an application's UI with
a spreadsheet that is synchronized with the application's data. When the
user manipulates the spreadsheet, the underlying data is modified and
the changes are propagated to the UI, and vice versa. We have
implemented this technique in a prototype browser extension called
Wildcard. Section~\ref{sec:demos} presents demos of using Wildcard to
augment real websites in useful ways, and
Section~\ref{sec:implementation} describes the implementation of the
extension.

@@

@@

Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
users to create ``spreadsheet-driven applications'' which use
spreadsheets as a backing data layer. Spreadsheet-driven
\emph{customization} applies this idea in a different context:
customizing existing software, rather than building new software from
scratch. Our technique does not require that the application actually be
backed by a spreadsheet; it merely uses the spreadsheet as an interface
for viewing and modifying the internal state of the application.

This approach requires extracting structured data from the user
interfaces of existing applications, but we hide the complexity of data
extraction from end users. Programmers write \emph{site adapters} which
use web scraping techniques to extract structured data from existing
applications and map them to the spreadsheet table. End users only
interact with the structured spreadsheet, providing a straightforward
customization experience. In Section~\ref{sec:related-work}, we explain
this architecture in more depth, and describe how Wildcard relates to
existing work on malleable software, web customization,
spreadsheet-driven applications, and web scraping.

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to explore how the
spreadsheet abstraction maps to real websites and customization needs.
Eventually we plan to release the tool publicly, in order to study real
applications, discover usability challenges, and to test the feasibility
of programmers building and maintaining site adapters.

\hypertarget{sec:demos}{%
\section{Demo: booking a trip with Wildcard}\label{sec:demos}}

To demonstrate the user experience of Wildcard, here's an example of
using it to help with booking a trip.{ These demos are best viewed as
videos in the online version of this paper
(\url{https://www.geoffreylitt.com/wildcard}).}

@@

In 2012, the travel site Airbnb removed the ability to sort
accommodation searches by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained that the change seemed hostile to users. ``It's so
frustrating!..What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

@@

Then, the user clicks on the price column header to sort the spreadsheet
and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}B)}. They also
filter to listings with a user rating above 4.5 (another feature missing
in the original Airbnb UI).

@@

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}D)}. Similarly, the
user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each listing
row, and the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved in the
browser for future sessions.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using Wildcard formulas, the user can integrate Airbnb
with Walkscore, an API that rates the walkability of any location on a
1-100 scale. When they type in a \texttt{walkscore} formula with the
latitude and longitude of the listing, it sets the walk score as the
value of the cell. Because the cell's contents are injected into the
page, the score also shows up in the UI{
(Figure~\ref{fig:airbnb-demo}C)}.

@@

It might seem that Wildcard is only useful on websites that display
lists of tabular data like search results. But in fact, the table
metaphor is flexible enough to represent many types of data. For
example, a flight search form on Expedia can be represented as a single
row, with a column corresponding to each input. {
Figure~\ref{fig:expedia-demo} shows an overview of augmenting the
Expedia site.}

In previous examples the table cells were read-only, since users cannot
change the name or price of an Airbnb listing. In this case, the cells
are writable, which means that changes in the table are reflected in the
form inputs. This becomes especially useful when combined with GUI
widgets that can edit the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates and then manually copying
them into the form. In Wildcard, the user can avoid this cumbersome
workflow by using a datepicker widget that includes the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}B)}. The user can
directly click on the right date, and it gets inserted into the
spreadsheet and the original form.

Here we've presented just a few use cases for spreadsheet-driven
customization. We think the spreadsheet model is flexible enough to
support a wide range of other useful modifications, while remaining
familiar and easy to use.

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{\{}
  \CommentTok{// Field metadata:}
\NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the field}
\NormalTok{  type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
\NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}

@@

Sometimes sophisticated scraping techniques are necessary to extract the
necessary data from the application. For example, when we have
prototyped mechanisms for site adapters to observe AJAX requests made by
the browser and extract data directly from JSON responses. This
mechanism was used to implement the Airbnb Walkscore example, since
latitude and longitude aren't shown in the Airbnb UI, but they are
available in AJAX responses. This technique seems promising because AJAX
responses tend to already contain data in a structured form, and web
applications increasingly load data using AJAX. Another technique we
might consider adding is the ability for site adapters to scrape data
across multiple pages for paginated lists of results (as explored in
\autocite{huynh2006}).

@@

\hypertarget{sec:design-principles}{%
\section{Design principles}\label{sec:design-principles}}

The idea of spreadsheet-driven customization is guided by several design
principles, inspired by prior work and our own experimentation. We think
these principles can also broadly inform the design of tools for end
user software customization.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
utilities, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
can get a lot of utility from deeply mastering a text editor and text
manipulation utilities like \texttt{cut} and \texttt{sed}, because these
tools can be applied to a huge variety of tasks. A user's preferred text
editor can even serve as an interactive input mechanism in shell
programs, e.g.~for editing git commit messages.

@@

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
which are UI elements that can be used in different contexts (for
example, a color picker that can be used in many different drawing
applications). It also relates to ideas of literacy in a medium. diSessa
notes that textual literacy rests on the fact that writing can be
adapted to many different genres and uses \autocite{disessa2000}; if
people needed to relearn reading and writing from scratch when switching
from essays to emails, the medium would lose most of its potency. Having
generic tools is especially important for software customization,
because the most common barrier to customizing software is not having
enough time \autocite{mackay1991}. It's more likely that people will
customize software regularly if they can reuse the same tools across
many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and can't necessarily naturally express
the data in every application. We plan to explore the limits of the
table abstraction further, by trying to build adapters for more sites
with varied data formats. We expect that many types of data can fit
easily into tables: lists of search results, news articles, and messages
can all naturally be seen as relations. For sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social
networks), it may prove more challenging to map internal data to the
table abstraction.

@@

One of the most interesting properties of spreadsheets is that users who
are only know a tiny sliver of their functionality (e.g., storing tables
of numbers or computing simple sums) can still use them in genuinely
valuable ways. The fact that useful tasks can be performed early on
supports the user's natural motivation to continue using the tool, and
to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront time investment before someone is able to
write a program that helps them achieve a useful task in their life.

As part of ensuring a low floor, we have focused on including genuinely
valuable features for novices. For example, a user can sort a table with
a single click, or simply type in some annotations. We would expect many
Wildcard users to start by using these simpler features before
potentially moving on to more sophisticated features like formulas.

@@

\begin{longtable}[]{@{}llc@{}}
\toprule
\endhead
& \emph{In-place} & \emph{Not in-place}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser JS console & forking open source
software\tabularnewline
\bottomrule
\end{longtable}

@@

Since we have still only built several site adapters and demos, it is
still too early to tell exactly how high the ceiling is for the
customizations that can be achieved with Wildcard. But we think that
with enough operators, the formula language could support a wide variety
of customizations---people have managed to use simple spreadsheet
formula languages to solve a surprisingly large range of problems. We
plan to explore this aspect further by trying to solve more real
problems with the system and observing where limitations emerge in
practice.

\hypertarget{build-for-multiple-tiers-of-users}{%
\subsection{Build for multiple tiers of
users}\label{build-for-multiple-tiers-of-users}}

@@

Another group to consider is the first party developers of the original
software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers, but if they were to
expose structured data in their web application clients, it would
eliminate the need for third party site adapters. We think there are
compelling reasons for first parties to consider doing this. Providing
Wildcard support would allow users to build extensions to fulfill their
own feature requests. It also would not necessarily require much effort,
since adding Wildcard support could be fairly straightforward for a
first-party that already has direct access to the structured data in the
page. There is also precedent for first parties implementing an official
client extension API: for several years, Google maintained an official
extension API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
  since then, third parties have continued to maintain stable Gmail
  extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

\hypertarget{sec:related-work}{%
\section{Related work}\label{sec:related-work}}

@@

An important design decision for tools in this space is how to deviate
from traditional spreadsheets like Microsoft Excel or Google Sheets.
Quilt and Glide use existing spreadsheet software as a backend,
providing maximum familiarity for users, and even compatibility with
existing spreadsheets. Gneiss has its own spreadsheet implementation
with additional features useful for building GUIs. Marmite provides a
live data view that resembles a spreadsheet, but programming is actually
done using a separate data flow pane rather than spreadsheet formulas.
(Marmite's approach led to some confusion in a user study, because users
expected behavior more similar to spreadsheets \autocite{wong2007}.)
Airtable deviates the furthest: although the user interface resembles a
spreadsheet, the underlying structure is a relational database with
typed columns. Wildcard's table is most similar to Airtable; the
structure of a relational table is most appropriate for most data in
websites, and we have not yet found a need for arbitrary untyped cells.

@@

\hypertarget{future-work}{%
\section{Future work}\label{future-work}}

@@

We plan to continue developing the Wildcard prototype and to eventually
deploy it as an open-source tool. To receive future updates on Wildcard
and notifications about a public release, \href{}{sign up for the email
newsletter}.

>>>>

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming tools, especially for live programming and end user programming. Contact him at \email{glitt@mit.edu}.}
\author{Daniel Jackson}
\authorinfo{is Professor in the Department of Electrical Engineering and Computer Science at MIT,  a MacVicar teaching fellow, and Associate Director of the Computer Science and Artificial Intelligence Laboratory. He leads the Software Design Group. Contact him at \email{dnj@csail.mit.edu}.}
\affiliation{Massachusetts Institute of Technology}

@@

We present the idea of \emph{spreadsheet-driven customization}: enabling
end users to customize existing applications using a live spreadsheet
view of the data inside the application. By manipulating the
spreadsheet, users can implement a variety of customizations without
doing any traditional programming.

We have implemented this technique in a prototype browser extension
called Wildcard. Through concrete examples, we demonstrate that Wildcard
can be used to implement useful customizations for real websites.

@@

In this paper, we present \emph{spreadsheet-driven customization}, a
technique that enables end users to customize software without doing any
traditional programming. The idea is to augment an application's UI with
a spreadsheet that is synchronized with the application's data. When the
user manipulates the spreadsheet, the underlying data is modified and
the changes are propagated to the UI, and vice versa.

We have implemented this technique in a prototype browser extension
called Wildcard. We've built demos of using Wildcard (shown in
Section~\ref{sec:demos}) which suggest that this paradigm can support
useful customizations, ranging from sorting lists of data to adding
whole new features to applications.

@@

Our approach requires extracting structured data from the user
interfaces of existing applications, but we hide the complexity of data
extraction from end users. Programmers write \emph{site adapters} which
use web scraping techniques to extract structured data from existing
applications and map them to the spreadsheet table. Our prototype
suggests that it is possible to implement site adapters for real
websites; Section~\ref{sec:implementation} describes some of the
techniques and challenges involved.

@@

Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
users to create ``spreadsheet-driven applications'' which use
spreadsheets as a backing data layer. Spreadsheet-driven
\emph{customization} applies this idea in a different context:
customizing existing software, rather than building new software from
scratch. Our technique does not require that the application actually be
backed by a spreadsheet; it merely uses the spreadsheet as an interface
for viewing and modifying the internal state of the application. In
Section~\ref{sec:related-work}, we describe how Wildcard relates to
existing work on spreadsheet-driven apps, as well as other areas like
web customization and web scraping.

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to explore how the
spreadsheet abstraction maps to real websites and customization needs.
Eventually we plan to release the tool publicly in order to study real
use cases, discover usability challenges, and to test the feasibility of
programmers building and maintaining site adapters.

\hypertarget{sec:demos}{%
\section{Demos}\label{sec:demos}}

Here are some examples of using Wildcard to customize websites in useful
ways.{ These demos are best viewed as videos in the online version of
this paper (\url{https://www.geoffreylitt.com/wildcard}).}

@@

In 2012, the travel site Airbnb removed the ability to sort
accommodation searches by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained that the change seemed hostile to users. ``It's so
frustrating! What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

@@

Then, the user clicks on the price column header to sort the spreadsheet
and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}, Note A)}.
They also filter to listings with a user rating above 4.5 (another
feature missing in the original Airbnb UI).

@@

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
the user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each listing
row, and the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
the browser and associated with the ID of the listing, so they will
appear in future sessions that display the same listing.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using a formula, the user can integrate Airbnb with
Walkscore, an API that rates the walkability of any location on a 1-100
scale. When the user uses the \texttt{walkscore} function with the
listing's latitude and longitude as inputs, it returns the walk score
for that location and shows it as the cell value. Because the cell's
contents are injected into the page, the score also appears in the UI{
(Figure~\ref{fig:airbnb-demo}, Note D)}.

\hypertarget{snoozing-todos}{%
\subsection{Snoozing todos}\label{snoozing-todos}}

\begin{figure}
\hypertarget{fig:todomvc-demo}{%
\centering
\includegraphics{media/todomvc-demo-300dpi.png}
\caption{Using Wildcard to add a ``snooze'' feature to the TodoMVC todo
list app}\label{fig:todomvc-demo}
}
\end{figure}

In addition to fetching data from other sources, Wildcard formulas can
also perform computations. In this example, the user wants to augment
the TodoMVC todo list app with a ``snooze'' feature, which will
temporarily hide a todo from the list until a certain date.{
Figure~\ref{fig:todomvc-demo} shows an overview of this customization.}

The user starts by opening the table view, which shows the text and
completed status of each todo, and live updates as the list changes.
They start the customization by adding a new \texttt{snoozeDate} column
to store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
Note A)}.

The next step is to hide snoozed todos. The user creates a new
\texttt{snoozed?} column, which uses a formula to compute whether a todo
has a snooze date in the future{ (Figure~\ref{fig:todomvc-demo}, Note
B)}. Then, they simply filter the table to hide the snoozed todos{
(Figure~\ref{fig:todomvc-demo}, Note C)}.

As time passes, the built-in \texttt{NOW()} function will return the
current date, and snoozed todos will automatically appear at the right
time.

Because this implementation of snoozing was built on top of the
spreadsheet abstraction, it is completely decoupled from this particular
todo list app. We envision that users could package and share these
types of customizations as generic browser extensions, which could then
be instantly applied to any site supported by Wildcard.

@@

It might seem that Wildcard is only useful on websites that display
lists of tabular data, but the table metaphor is flexible enough to
represent many types of data. For example, a flight search form on
Expedia can be represented as a single row, with a column corresponding
to each input{ (Figure~\ref{fig:expedia-demo}, Note A)}.

In some of the previous examples, the table cells were read-only
(because users can't, for example, change the name or price of an Airbnb
listing). In this case, the cells are writable, which means that changes
in the table are reflected in the form inputs. This becomes especially
useful when combined with GUI widgets that can edit the value of a table
cell.

Filling in dates for a flight search often requires a cumbersome
workflow: open up a separate calendar app, find the dates for the trip,
and then carefully copy them into the form. In Wildcard, the user can
avoid this by using a datepicker widget that shows the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
directly click on the correct date, and it gets inserted into both the
spreadsheet and the original form.

In this section we've presented just a few use cases for
spreadsheet-driven customization, to hint at the possibilities of the
paradigm. We think the spreadsheet model is flexible enough to support a
wide range of other useful modifications, while remaining familiar and
easy to use.

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{\{}
  \CommentTok{// Field metadata:}
\NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
\NormalTok{  type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
\NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}

@@

Sometimes sophisticated scraping techniques are necessary to extract the
necessary data from the application. For example, we have prototyped a
mechanism for site adapters to observe AJAX requests made by the browser
and extract data directly from JSON responses. This mechanism was used
to implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
This technique seems promising because AJAX responses tend to already
contain data in a structured form, and web applications increasingly
load data using AJAX. Another technique we might consider adding is the
ability for site adapters to scrape data across multiple pages for
paginated lists of results (as explored in \autocite{huynh2006}).

@@

\hypertarget{sec:design-principles}{%
\section{Design Principles}\label{sec:design-principles}}

The idea of spreadsheet-driven customization is guided by several design
principles, inspired by prior work and our own experimentation. We think
these principles might also broadly inform the design of tools for end
user software customization.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
programs, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
benefit from deeply mastering a text editor and other text manipulation
tools, since they can be applied to a huge variety of tasks. A user's
preferred text editor can even serve as an interactive input mechanism
in shell programs, e.g.~for editing git commit messages.

@@

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
which are UI elements that can be used in different contexts (for
example, a color picker that can be used in many different drawing
applications). It also relates to ideas of literacy in a medium. DiSessa
notes that textual literacy rests on the fact that writing can be
adapted to many different genres and uses \autocite{disessa2000}; if
people needed to relearn reading and writing from scratch when switching
from essays to emails, the medium would lose most of its potency. Having
generic tools is especially important for software customization,
because the most common barrier to customizing software is not having
enough time \autocite{mackay1991}. It's more likely that people will
customize software regularly if they can reuse the same tools across
many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and may not naturally express the data
in every application. We plan to explore the limits of the table
abstraction further, by trying to build adapters for more sites with
varied data formats. We expect that many types of data can fit easily
into tables: lists of search results, news articles, and messages can
all naturally be seen as relations. For sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social
networks), it may prove more challenging to map internal data to the
table abstraction.

@@

One of the most interesting properties of spreadsheets is that users
familiar with only a tiny sliver of their functionality (e.g., storing
tables of numbers and computing simple sums) can still use them in
genuinely valuable ways. This supports the user's natural motivation to
continue using the tool, and to eventually learn its more powerful
features if needed \autocite{nardi1991}. In contrast, many traditional
programming systems require an enormous upfront time investment before
someone is able to write a program that helps them achieve a useful task
in their life.

As part of ensuring a low floor, we have focused on including features
that have immediate value for novices. For example, a user can sort a
table with a single click, or simply type in some annotations. We would
expect many Wildcard users to start by using these simpler features
before potentially moving on to more sophisticated features like
formulas.

@@

\begin{longtable}[]{@{}llc@{}}
\toprule
\endhead
& \emph{In-place} & \emph{Not in-place}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser JS console & using Web APIs in
Python\tabularnewline
\bottomrule
\end{longtable}

@@

Since we have still only built a few site adapters and demos, it is
still too early to tell exactly how high the ceiling is for the
customizations that can be achieved with Wildcard. But we think that
with enough built-in formulas (including standard spreadsheet formulas
and formulas for fetching data from other sources), the language could
support a wide variety of customizations. We plan to explore this aspect
further by trying to solve more real problems with the system and
observing where limitations emerge in practice.

\hypertarget{build-for-an-ecosystem}{%
\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}

@@

The first party developers of the original software can also play a role
in enabling customization. Spreadsheet-driven customization does not
depend on cooperation from first-party website developers, but if they
were to expose structured data in their web application clients, it
would eliminate the need for third party site adapters. We think there
are compelling reasons for first parties to consider doing this.
Providing Wildcard support would allow users to build extensions to
fulfill their own feature requests. It also would not necessarily
require much effort, since adding Wildcard support could be fairly
straightforward for a first-party that already has direct access to the
structured data in the page. There is also precedent for first parties
implementing an official client extension API: for several years, Google
maintained an official extension API in Gmail for Greasemonkey scripts
to use.\footnote{Incidentally, since then, third parties have continued
  to maintain stable Gmail extension APIs used by many browser
  extensions \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

\hypertarget{sec:related-work}{%
\section{Related Work}\label{sec:related-work}}

@@

An important design decision for tools in this space is how far to
deviate from traditional spreadsheets like Microsoft Excel or Google
Sheets. Quilt and Glide use existing spreadsheet software as a backend,
providing maximum familiarity for users, and even compatibility with
existing spreadsheets. Gneiss has its own spreadsheet implementation
with additional features useful for building GUIs. Marmite provides a
live data view that resembles a spreadsheet, but programming is actually
done using a separate data flow pane rather than spreadsheet formulas.
(Marmite's approach led to some confusion in a user study, because users
expected behavior more similar to spreadsheets \autocite{wong2007}.)
Airtable deviates the furthest: although the user interface resembles a
spreadsheet, the underlying structure is a relational database with
typed columns. Wildcard's table is most similar to Airtable; the
structure of a relational table is most appropriate for most data in
websites, and we have not yet found a need for arbitrary untyped cells.

@@

\hypertarget{future-work}{%
\section{Future Work}\label{future-work}}

@@

We plan to continue developing the Wildcard prototype and to eventually
deploy it as an open-source tool. To receive future updates on Wildcard
and notifications about a public release,
\href{https://geoffreylitt.substack.com/welcome}{sign up for the email
newsletter}.
